MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-08-28 latex/paper/sections/12-future_work.tex
review kraftwerk

Unlike SA, which primarily relies on heuristics and probabilistic moves, AP begins by contextualizing the problem as a collection of mathematical expressions compatible with convex optimization solvers.
As briefly introduced in Section~\ref{subsec:netlist}, a circuit netlist can be naturally modeled as a hypergraph
\begin{equation}
    G_{H} = (V_{H}, E_{H})
    \label{equ:hypergraph}
\end{equation}
where \(V_{H}\) is the set of vertices and \(E_{H}\) is the set of hyperedges, each of which may connect more than two vertices. 
In FPGA context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s connecting them.

This hypergraph can be reduced into a graph using one of several net models:
\begin{itemize}
    \item \textbf{Star model:} classically, introduce a virtual star node and connect it to all pins of the net. However, in electronics, we can simply use the voltage source node as the star node and avoid adding virtual nodes.
    \item \textbf{Clique model:} connect all pins of the net to each other (quadratic growth in edges). 
    \item \textbf{Bound2Bound (B2B) model:} connect only the pins at the geometric boundary (min/max in $x$ or $y$) to each other, and connect boundary pins to all interior pins. With inverse-distance weights, this model matches HPWL exactly at the point of weight calculation.
\end{itemize}

Since HPWL is direction-agnostic, directionality of FPGA nets (source/sinks) is typically ignored in reduction. 
Figure \ref{fig:net_model} shows an example of hyperedge reduction using the star and clique models.

@@

\begin{equation}
    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
    \label{equ:Manhattan}
\end{equation}

\begin{equation}
    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
    \label{equ:Euclidian}
\end{equation}

Here, module placements in the $x$ and $y$ directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \), where $x_n$ and $y_n$ correspond to the coordinates of the $n$th module instance.
In both formulations, the objective function is separable with $x$ and $y$.
The weight \(w_{i,j}\) is the weight of the 2-pin edge connecting nodes $i$ and $j$. 
Its value depends on both the net model and the objective:

\begin{itemize}
    \item \textbf{Star/Clique models with a quadratic (squared-distance) objective:}  
    Use constant edge weights. In the star model, each reduced 2-pin edge is given unit weight $w_{ij}=1$.  
    In the clique model, each pin–pin edge is given weight $w_{ij}=1/(p-1)$, where $p$ is the pin count of the original hyperedge from which this edge was produced.  
    This scaling prevents large-fanout nets from dominating the cost and makes star and clique models equivalent in the quadratic formulation.

@@

    \item \textbf{Bound2Bound model with a linear HPWL objective:}  
    Connect only boundary–boundary and boundary–interior pin pairs, and assign weights as inverse distances:  
    \begin{equation}
        w_{i,j} = \frac{2}{(p-1)\,|x_i - x_j|}.
        \label{equ:weight_linearized}
    \end{equation}
    With this assignment, the quadratic cost of a net equals its HPWL (in each dimension) at the evaluation point, making the Bound2Bound model HPWL-accurate.
\end{itemize}

@@

For ease of implementation, it would be best to first implement the squared objective before exploring the linear objective.
We can use the FPGA placer HeAP (2012) \cite{AP_2012}, which uses the squared objective along with the B2B net model, as an instructive reference.
We can cast the squared formulation into quadratic form. 
In the x-direction this can be written conveniently as
\begin{equation}
    \boldsymbol{\Phi} (\boldsymbol{x}) 
    = \tfrac{1}{2}\,\boldsymbol{x}^T \boldsymbol{Q_x}\,\boldsymbol{x} + \boldsymbol{c_x}^T \boldsymbol{x} + \text{const},
    \label{equ:quadratic}
\end{equation}
where $\boldsymbol{Q_x}$ encodes 2-pin connections between movable cells and $\boldsymbol{c_x}$ accounts for fixed terminals.  
Solving $\boldsymbol{Q_x}\boldsymbol{x} = -\boldsymbol{c_x}$ yields the optimal continuous $x$ coordinates, and the same applies independently for $y$.
This quadratic form lends itself nicely to some OTS quadratic solvers.

>>>>

Unlike SA, which primarily relies on heuristics and probabilistic moves, AP begins by contextualizing the problem as a collection of mathematical expressions compatible with convex optimization solvers.
As briefly introduced in Section~\ref{subsec:netlist}, a circuit netlist can be naturally modeled as a hypergraph
\begin{equation}
    G_{H} = (V_{H}, E_{H})
    \label{equ:hypergraph}
\end{equation}
where \(V_{H}\) is the set of vertices and \(E_{H}\) is the set of hyperedges, each of which may connect more than two vertices. 
In RapidWright context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s connecting them.
More accurately, $V_H$ corresponds to the set of pins (\texttt{SitePinInst}, or SPI for short), not the \texttt{SiteInst}s themselves.
The distinction is usually only important in VLSI context, when the area of a module or macro instance is large enough that the distance between its pins becomes significant \cite{kraftwerk2}.
However, in the context of FPGAs where \texttt{Site}s are relatively uniform size and whose areas are small compared to the area of the \texttt{device}, all \texttt{SitePinInst}s belonging to the same \texttt{SiteInst} can share the same location coordinates.

(SHOW THE NOTATION)

Modules 
$m \in M$

Pins
$p \in P$

Nets (hyperedges)
$n \in N$

$e \in {\mathcal{E}}_n$

$e = (p, q)$

This hypergraph can be reduced into a graph using one of several net models:
\begin{itemize}
    \item \textbf{Star model:} classically, introduce a virtual star node and connect it to all pins of the net. However, in electronics, we can simply use the voltage source node as the star pipipinnnnode and avoid adding virtual nodes.
    \item \textbf{Clique model:} connect all pins of the net to each other (quadratic growth in edges). 
    \item \textbf{Bound2Bound (B2B) model:} connect only the pins at the geometric boundary (min/max in $x$ or $y$) to each other, and connect boundary pins to all interior pins. With inverse-distance weights, this model matches HPWL exactly at the point of weight calculation.
\end{itemize}

Since HPWL is direction-agnostic, directionality of FPGA nets (source/sinks) is typically ignored in reduction. 
Figure \ref{fig:net_model} shows an example of hyperedge reduction using the star and clique models while \ref{fig:bound2bound} shows the B2B reduction.

@@

\begin{equation}
    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{ij} \left( |x_i - x_j| + |y_i - y_j| \right)
    \label{equ:Manhattan}
\end{equation}

\begin{equation}
    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{ij} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
    \label{equ:Euclidian}
\end{equation}

Here, module placements in the $x$ and $y$ directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \), where $x_n$ and $y_n$ correspond to the coordinates of the $n$th module instance.
In both formulations, the objective function is separable with $x$ and $y$.
The weight \(w_{ij}\) is the weight of the edge connecting pins $i$ and $j$. 
Its value depends on both the selected net model and objective:

\begin{itemize}
    \item \textbf{Star/Clique models with a quadratic (squared-distance) objective:}  
    Use constant edge weights. In the star model, each pin-pin edge is given unit weight $w_{ij}=1$.  
    In the clique model, each pin–pin edge is given weight $w_{ij}=1/(p-1)$, where $p$ is the pin count of the original hyperedge from which this edge was produced.  
    This scaling prevents large-fanout nets from dominating the cost and makes star and clique models equivalent in the quadratic formulation.

@@

    \item \textbf{Bound2Bound model with a linear HPWL objective:}  
    Connect only boundary–boundary and boundary–interior pin pairs, and assign weights as inverse distances:  
    \begin{equation}
        w_{ij} = \frac{2}{(p-1)\,|x_i - x_j|}.
        \label{equ:weight_linearized}
    \end{equation}
    With this assignment, the quadratic cost of a net equals its HPWL (in each dimension) at the evaluation point, making the Bound2Bound model HPWL-accurate.
\end{itemize}

@@

For ease of implementation, it would be best to first implement the squared objective before exploring the linear objective.
We can use the FPGA placer HeAP (2012) \cite{AP_2012}, which uses the squared objective along with the B2B net model, as an instructive reference.

We can cast the squared formulation (\ref{equ:Euclidian}) into standard quadratic form. 
The derivation is shown in Kraftwerk2 (2008) \cite{kraftwerk2}.
Take the original squared objective in the $x$ direction (\ref{equ:sq_obj_x}) and expand each term (\ref{equ:w_expand}).
\begin{equation}
    {\boldsymbol{\Phi}}_x (\boldsymbol{x}) = \sum_{i, j} w_{ij} \left( x_i - x_j \right)^2
    \label{equ:sq_obj_x}
\end{equation}

\begin{equation}
    w_{ij}(x_i - x_j)^2 = w_{ij} \left(x_i^2 - 2 x_i x_j + x_j^2 \right)
    \label{equ:w_expand}
\end{equation}

In the x-direction this can be written as
\begin{equation}
    \boldsymbol{\Phi} (\boldsymbol{x}) 
    = \tfrac{1}{2}\,\boldsymbol{x}^T \boldsymbol{Q}_x\,\boldsymbol{x} + \boldsymbol{c}_x^T \boldsymbol{x} + \text{const},
    \label{equ:quadratic}
\end{equation}
where $\boldsymbol{Q_x}$ encodes 2-pin connections between movable cells and $\boldsymbol{c_x}$ accounts for fixed terminals.  
Solving $\boldsymbol{Q_x}\boldsymbol{x} = -\boldsymbol{c_x}$ yields the optimal continuous $x$ coordinates, and the same applies independently for $y$.
This quadratic form lends itself nicely to some OTS quadratic solvers.
